---
title: "Clusters"
author: "Ethan Welty"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Clusters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Falling Fruit uses clusters to summarize the number and location of types available over large areas. The clusters are computed using a very fast algorithm described below.

## Convert to Web Mercator

Coordinates are stored in the database as standard WGS 84 longitude and latitude in decimal degrees (EPSG:4326).

```{r}
locations <- data.table::data.table(
  lng = c(-90, 90, -90, 90),
  lat = c(-45, 45, 45, -45)
)
```

Because web mapping applications use the Web Mercator projection, we first convert our coordinates to Web Mercator (EPSG:3857).

```{r}
locations[, c("x", "y") := as.data.frame(
  fruitr::lnglat_to_xy(cbind(lng, lat))
)]
```
```{r, echo=FALSE}
locations
```

## Quadtree the Earth

In the Web Mercator projection, if you cut off the poles (at roughly +/- 85° latitude), the surface of the Earth is conveniently represented by a square (`zoom = 0`). This square Earth can be divided into four smaller squares (`zoom = 1`), each of which can be divided into four even smaller squares, and so on. At any given zoom level in this [quadtree](https://en.wikipedia.org/wiki/Quadtree), the Earth is divided into 4<sup>zoom</sup> equal-sized squares.

The squares (grid "cells") at each zoom level can be numbered with a simple x and y integer index starting from the cell (0, 0) at the bottom left corner (near -180° longitude, -85° latitude). For speed, we compute grid cell indices for only the highest supported zoom level, then build the lower zoom levels from these. Falling Fruit computes clusters to `zoom = 13`, but for the sake of example, let's use `zoom = 1`.

```{r}
locations[, c("xi", "yi", "zoom") := as.data.frame(
  fruitr::xy_to_gridcells(cbind(x, y), zoom = 1)
)]
```
```{r, echo=FALSE}
locations
```

With grid cell indices assigned to each location, we can now calculate the total count and center of mass of all locations in each grid cell.

```{r}
clusters <- locations[,.(x = mean(x), y = mean(y), count = .N), by = .(xi, yi, zoom)]
```
```{r, echo=FALSE}
clusters
```

## Geohash the parents

But how do we generate the lower zoom level clusters? As it turns out, from the grid cell indices we can construct binary [geohashes](https://en.wikipedia.org/wiki/Geohash) which "magically" contain the indices of all parent cells!

At `zoom = 0`, one cell encompasses the entire Earth. Counting from zero, that cell is number zero. In binary, the number zero (0) is "00", which is also the x and y grid indices for that cell: (0, 0).

At `zoom = 1`, the four cells are:

 - 0: "00" or (0, 0) (bottom left)
 - 1: "01" or (0, 1) (top left)
 - 2: "10" or (1, 0) (bottom right)
 - 3: "11" or (1, 1) (top right)
 
To construct our geohashes, we convert the x and y grid indices to binary and interleave the bits. For example, cell (1, 0) - which is ("01", "00") in binary - becomes "00 10". Therefore, cell "00 10" (1, 0) at `zoom = 1` is the child of cell "00" (0, 0) at `zoom = 0`. At `zoom = 2`, the cell (2, 1) is ("010", "001") in binary (left-padded with zeros to account for the higher zoom level). Interleaved, this becomes "00 10 01", from which we know that cell "00 10 01" (2, 1) at `zoom = 2` is the child of cell "00 10" (1, 0) at `zoom = 1` and cell "00" (0, 0) at `zoom = 0`.

In practice, we compute the geohashes of our base clusters, then replicate the cluster for each of its parents by expanding its geohash.

```{r}
clusters[, geohash := fruitr::gridcells_to_geohashes(cbind(xi, yi, zoom))]
geohashes <- clusters[, fruitr::expand_geohashes(geohash)]
clusters <- clusters[rep(1:.N, nchar(geohash) / 2)]
clusters[, geohash := geohashes]
```
```{r, echo=FALSE}
clusters
```

The total count and center of mass of each parent cluster can then be calculated from all rows with the same geohash.

```{r}
clusters <- clusters[, 
    .(x = weighted.mean(x, count), y = weighted.mean(y, count), count = sum(count)),
    by = geohash
]
```
```{r, echo=FALSE}
clusters
```

## Finalize clusters

The zoom level of each cluster is a simple function of the length of it's geohash (`nchar(geohash) / 2 - 1).

```{r}
clusters[, zoom := fruitr::geohashes_to_zoom(geohash)]
```
```{r, echo=FALSE}
clusters
```

Finally, the centers of mass computed in Web Mercator are converted to WGS 84 longitude and latitude.

```{r}
clusters[, c("lng", "lat") := as.data.frame(
  fruitr::xy_to_lnglat(cbind(x, y))
)]
```
```{r, echo=FALSE}
clusters
```

Which we can check against our original locations.

```{r, echo=FALSE}
locations[, c("lng", "lat")]
```

## Update clusters

As locations are added or removed, the clusters need to be updated. The geohashes of all cells containing the location are computed (with the process described above) and used to quickly lookup and increment (or decrement) all impacted clusters.
